IML Term project paper

### 2022-12-11 paper.rmd
Generative models.

```{r}
npf<-read.csv("C:\\Users\\pyrys\\Downloads\\Koulujuttuja\\IntroMachineLearning\\npf_train.csv")
npf$class2<-npf$class4!="nonevent"
print(paste0("Relative amount of observations in class nonevent: ",mean(npf$class4=="nonevent")))
print(paste0("Relative amount of observations in class Ia: ",mean(npf$class4=="Ia")))
print(paste0("Relative amount of observations in class Ib: ",mean(npf$class4=="Ib")))
print(paste0("Relative amount of observations in class II: ",mean(npf$class4=="II")))
```

@@

%Mitä teemme tälle, haluammeko viitata tähän visualisointien ja nimien suhteen vai onko ideaalia, että teemme omat visualisointimme, jotka esittelemme tekstissä?
We also familiarized ourselves with the data through the [smear.avaa.scs.fi](https://smear.avaa.csc.fi/) webpage that offers further visualizations details about the data and the measurement site, and with the variable names and details available at [wiki.helsinki.fi](https://wiki.helsinki.fi/pages/viewpage.action?pageId=243959901).

@@

(The data is over non-consecutive days, ok, but are there any significant gaps e.g. a month?)

@@

```{r original correlation plot, echo=FALSE, fig.dim=c(7, 7), message=FALSE}
library(corrplot)
# Original correlations
cmData <- cor(npfData)
corrplot(cmData, order = "FPC", tl.cex = 0.3, tl.col = "black")
``` 
Many classifiers are affected by correlation and colinearity between variables. As expected, we found that many of the variables describing the same phenomenon are correlated, as are variables related to radiation (see figure below).

The initial data analysis showed that many of the parameters are correlated. For many classifiers, highly correlated parameters are problematic, and for this reason we set out to drop some of the most highly correlated columns from the data set. We know that many of the parameters are measurements of the same thing at different heights. For example, water vapour concentration (H20) has been measured at heights 4.2m, 8.4m, 16.8m, 33.6m, 50.4m and 67.2m. The correlation coefficients between the daily means at different heights are essentially 1 (see table 1).

Hence we decided to discard most of the measurements of H2O and other variables for which we had multiple correlating measurements, both the means and the standard deviations. We ended up keeping the measurements measured at 16.8 meters as for some parameters, the measurements were taken only at that height. We found that after this, there are 38 unique columns. The figure below describes the correlations after omitting all but the measurements taken at the selected height.

@@

@@

### Support Vector Classifiers

The idea of the binary support vector classifier is based on the notion of finding within a $p$-dimensional problem a $(p-1)$-dimensional hyperplane $\lambda_0+\sum\lambda_i x_i=0$ to divide the set of explanatory variables. Now, for this hyperplane, we can divide our observations into two groups, those values for which $\lambda_0+\sum_{i=1}^m\lambda_j x_i>0$ and those values for which $\lambda_0+\sum\lambda_i x_i<0$. Thus, this can function as a model through which we can predict binary variables such as our variable of interest.

However, defining such hyperplanes can be slightly difficult as all values might not be separable by a hyperplane or the hyperplane that separates values can overfit to training data. Thus, we define a soft-margin classifier, which has the benefits of letting some values be misclassified or be closer to the hyperplane than a strict maximal margin classifier would define, which leads to a more robust model that fits most of the training dataset better.

In order to define this, we need to find the optimal hyperplane to separate the observations. This optimization problem is an optimization problem of $M$ in relation to coefficients $\lambda_i$ and slack variables $\epsilon$ within inequation $y_i(\lambda_0+\sum_{j=1}^m\lambda_jx_{ij})\ge M(1-\epsilon_i)$ with limitations $\sum_{j=1}^{m}\lambda_j^2=1$, $\epsilon_i\ge0$ and $\sum_{i=1}^n\epsilon_i\le C$ for tuning parameter $C$.

As in our optimization the data only depends on the inner products of the explanatory variables and the values of the variable of interest, we can use different generalizations of inner products to gain different kinds of models. Thus with different choices of kernels we can model different kinds of data.

Other benefits benefits of using support vector classifiers include that they have been proven to behave well in several different applications and that it has been shown that on average support vector classifiers perform better than logistic regression models in problems where the classes are well-separated. However, the model can vary quite notably based on the tuning parameter $C$ as this parameter affects the amount of observations that violate the margin and thus increases or decreases the amount of support vectors that influence the model.

For our model, we chose to run a 10-fold cross validation on a set of tuning parameters $C$ to assess which kernel performed best and chose to model the value based on that. In these preliminary assessments, we assessed the radial kernel had the best performance of the ones available in the R package e1071. For this kernel, we ran loo-cross validation to assess the value of tuning parameter $C$ and calculated for said tuning parameter loo-cross-validated accuracy and perplexity.

>>>>

```{r}
npf<-read.csv("npf_train.csv")
npf$class2<-npf$class4!="nonevent"
print(paste0("Relative amount of observations in class nonevent: ",mean(npf$class4=="nonevent")))
print(paste0("Relative amount of observations in class Ia: ",mean(npf$class4=="Ia")))
print(paste0("Relative amount of observations in class Ib: ",mean(npf$class4=="Ib")))
print(paste0("Relative amount of observations in class II: ",mean(npf$class4=="II")))
```

@@

%Mitä teemme tälle, haluammeko viitata tähän visualisointien ja nimien suhteen vai onko ideaalia, että teemme omat visualisointimme, jotka esittelemme tekstissä?
% (Sara) Ainakin jos tehdään omat visualisoinnit, mielestäni tämän voi poistaa. Visualisointien kanssa kuitenkin miettisin, että mitä todella halutaan näyttää/kertoa, ja tukeeko kyseinen tai kyseiset visualisoinnit tätä. Eli jos visualisointeja halutaan, niin mitä ja miksi, ja miten ne liittyvät luokitteluun ja partikkeleiden muodostumiseen tai muuten tukevat projektityön päämääriä. 
We also familiarized ourselves with the data through the [smear.avaa.scs.fi](https://smear.avaa.csc.fi/) webpage that offers further visualizations details about the data and the measurement site, and with the variable names and details available at [wiki.helsinki.fi](https://wiki.helsinki.fi/pages/viewpage.action?pageId=243959901).

@@

@@

The correlation between the predictors and the value to be predicted is a very natural starting point when it comes to correlation analysis. It is also possible that the predictors themselves are correlated. Many classifiers are affected by high correlation, i.e. *collinearity* between variables. In regression models, for example, it can be difficult to differentiate between the individual effects of collinear variables on the response.

```{r original correlation plot, echo=FALSE, fig.dim=c(7, 7), message=FALSE}
library(corrplot)
# Original correlations
cmData <- cor(npfData)
corrplot(cmData, order = "FPC", tl.cex = 0.3, tl.col = "black")
```

As expected, the correlation plot shows that many of the variables describing the same phenomenon at different heights are correlated, as are variables related to radiation. For many classifiers, highly correlated parameters are problematic, and for this reason we would like to drop some of the most highly correlated columns from the dataset.

We know that many of the parameters are measurements of the same thing at different heights. For example, water vapour concentration (H20) has been measured at heights 4.2m, 8.4m, 16.8m, 33.6m, 50.4m and 67.2m. As can be seen from the table below, the correlation coefficients between the daily means of water vapour concentration at different heights are essentially 1.

@@

Hence we discard most of the measurements of H2O and other variables for which we have multiple highly correlating measurements, both the means and the standard deviations. We keep the parameters measured at 16.8 meters as for some parameters, the measurements have only been measured at that height. This rather crude method of discarding data leaves us with 38 unique columns. The figure below describes the correlation coefficients between the remaining parameters.

@@

## Generative models

In logistic regression models, we directly model $p(Y|X)$, i.e. the conditional distribution of the response $Y$ given the predictors $X$. Generative models offer a less direct approach, where the focus is in modeling $p(X|Y)$ and using these estimates to estimate $p(Y|X)$ for each possible class $k$ with Bayes' theorem:
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X|Y=k)}{\sum_i \pi_i p(X|Y=i)}.
\end{equation*}
Here $pi_k$ is an estimate of the prior probability that a random observation comes from the $k$th class, in our case computed as the fraction of the observations in the training dataset that are in the $k$th class.

While the general approach is the same for all generative models, they differ in how they estimate $p(X|Y)$. In a dataset with $p$ predictors, estimating $p(X|Y)$ amounts to estimating a $p$-dimensional density function for an observation in the $k$th class. The task is challenging, as we must consider the distribution of each predictor on its own *and* the joint distribution of the predictors. Different models make different assumptions that mitigate the difficulty.

### Linear Discriminant Analysis (LDA)

A linear discriminant analysis classifier assumes that all $p$ predictors $X = (X_1, \dots, X_p)$ are drawn from a multivariate Gaussian distribution. This means that each predictor follows a normal distribution $N(\mu_k, \Sigma)$ where $\mu_k$ is the class-specific mean and $\operatorname{Cov}(X) = \Sigma$ the covariance matrix of $X$, and there is some correlation between each pair of predictors. The estimate of $p(X|Y)$ is
\begin{equation*}
p(X|Y) = \frac{1}{(2\pi)^{p/2}|\Sigma|^{1/2}} \exp\left( -\frac{1}{2}(x-\mu_k)^T\Sigma^{-1}(x-\mu_k) \right)
\end{equation*}
meaning that the classifier assigns an observation $X$ to to the class for which
\begin{equation*}
\delta_k(x) = x^T\Sigma^{-1}\mu_k-\frac{1}{2}\mu_k^T\Sigma^{-1}\mu_k + \log \pi_k
\end{equation*}
is maximized. 
The covariance matrix being the same for all classes is a defining feature of linear discriminant analysis. The assumption makes the model linear and reduces the number of parameters to estimate. This reduces the flexibility and hence lowers the variance of the model. Lower variance can mean that the model performs well, but if the assumption is not reflected in the data, bias can be high and overall performance suffers due to the bias-variance trade-off.

For linear discriminant analysis to work, it is important to have enough observations with regards to the number of predictors, as performance suffers greatly as the number of predictors approaches the number of observations.

In our modeling, we used the function *lda* from the library *MASS*. We tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

### Quadratic Disriminant Analysis

The assumption that the covariance matrix is shared for all classes is quite strict. Allowing each class to have it's own covariance matrix brings us to quadratic discriminant analysis (QDA). The observations are still assumed to be drawn from a multivariate Gaussian distrbution with a class-specific mean vector, so otherwise the assumptions remain the same as in LDA. Each predictor now follows a normal distribution $N(\mu_k, \Sigma_k)$. The expression to be maximized is 
\begin{align*}
\delta_k(x) = x^T\Sigma_k^{-1}\mu_k-\frac{1}{2}x^T\Sigma_k^{-1}x-\frac{1}{2}\mu_k^T\Sigma_k^{-1}\mu_k -\frac{1}{2} \log |\Sigma_k| + \log \pi_k.
\end{align*}
Generally, quadratic discriminant analysis is expected to perform better that linear discriminant analysis on large data sets, but as in linear discriminant analysis, the number of predictors must be small enough to produce decent results. Quadratic discriminant analysis is also recommended when the assumption that all classes share a covariance matrix is clearly unfounded.

In our approach, we used the *qda* function from the *MASS* library, and as with linear discriminant analysis, tested the performance of the model on both the original dataset and the dataset with observations measured at multiples only observed at 16.8 meters.

### Naive Bayes

A Naive Bayes classifier makes no assumptions on the distribution of the observations, and instead assumes that within the *k*th class, the *p* predictors are independent. In other words, the assumption is that there is no association between the predictors, and therefore no joint distribution to consider.

The assumption of independence between predictors is often not realistic, but the model can perform well nonetheless. This is especially true for small data sets, since estimating a joint distribution requires a large amount of data.
